---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# irps: Immune Response Potential Scoring

[![R-CMD-check](https://github.com/maynard/irps/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/maynard/irps/actions/workflows/R-CMD-check.yaml)

The **irps** R package computes the **Immune Response Potential Score (IRPS)** from proteomics or transcriptomics data to evaluate PD-1/PD-L1 immunotherapy response potential. It provides:

- **18 curated immune gene sets** spanning antigen presentation, immune cell populations, immunosuppression, and inflammatory pathways
- **Single-sample GSEA** for per-sample immune microenvironment profiling
- **IRPS composite score**: a single scalar balancing favorable vs unfavorable immune signals
- **7-dimension portrait decomposition** for interpretable immune characterization
- **Immune phenotype classification** (Inflamed / Excluded / Desert), adapted from Chen & Mellman (2017) for body-fluid proteomics
- **MSigDB Hallmark validation** for external benchmarking of custom gene sets
- **Comprehensive visualization suite**: boxplots, heatmaps, radar charts, dimension panels

## Installation

```r
# Install from GitHub
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("maynard/irps")
```

## Quick Start

```{r, eval = FALSE}
library(irps)

# Get the built-in immune gene sets
gs <- get_immune_gene_sets()
names(gs)
# [1] "01_Antigen_Presentation"  "02_MHC_I_Core"           ...

# Example: proteomics data
# expr_mat: proteins × samples matrix (normalized, e.g., median-centered log2)
# gene_map: maps gene symbols to protein IDs in your data
head(gene_map)
#   Gene Protein
# 1  B2M  P61769
# 2 HLA-A P04439
# ...

result <- compute_irps(
  expr_mat  = expr,
  gene_map  = gene_map,
  platform  = "proteomics"
)

# Inspect results
print(result)
summary(result)

# Visualize
plot(result, type = "boxplot")
plot(result, type = "heatmap")
plot(result, type = "radar_favorable")
```

## Core Functions

| Function | Description |
|----------|-------------|
| `compute_irps()` | Main function: run ssGSEA and compute IRPS |
| `decompose_portrait()` | Decompose IRPS into 7 interpretable dimensions |
| `classify_phenotype()` | Classify samples as Inflamed/Excluded/Desert |
| `validate_hallmark()` | Validate custom gene sets against MSigDB Hallmarks |
| `get_immune_gene_sets()` | Return the 18 built-in immune gene sets |
| `map_genes_to_features()` | Map gene symbols to feature IDs in expression data |

## Visualization Functions

| Function | Output |
|----------|--------|
| `plot_irps_boxplot()` | Boxplot of IRPS scores by group |
| `plot_irps_heatmap()` | Clustered heatmap of ssGSEA scores |
| `plot_irps_radar()` | Radar chart of gene set scores |
| `plot_portrait_panel()` | 6-dimension faceted boxplot panel |
| `plot_phenotype_bar()` | Stacked bar chart of immune phenotypes |
| `plot_irps_scatter()` | IRPS vs individual biomarker scatter |

## IRPS Methodology

**IRPS** = mean(5 Favorable gene sets) − mean(3 Unfavorable gene sets)

**Favorable (higher = better for immunotherapy):**
1. Antigen Presentation (MHC-I/B2M/TAP pathway)
2. MHC-I Core
3. IFN-γ Response
4. Cytotoxic Potential (surrogate)
5. Chemokine/Inflammatory

**Unfavorable (higher = worse for immunotherapy):**
1. Immunosuppression (TGFB, CD47, S100A8/A9, collagens)
2. Stromal/CAF barrier
3. MDSC/Neutrophil infiltration

The score integrates these 8 dimensions into a single value where higher IRPS indicates a more favorable immune microenvironment for PD-1/PD-L1 blockade therapy.

## References

- Sharma P, et al. Primary, Adaptive, and Acquired Resistance to Cancer Immunotherapy. *Cell*. 2017;168(4):707-723.
- Kalbasi A, Ribas A. Tumour-intrinsic resistance to immune checkpoint blockade. *Nat Rev Immunol*. 2020;20(1):25-39.
- Chen DS, Mellman I. Elements of cancer immunity and the cancer-immune set point. *Nature*. 2017;541(7637):321-330.
- Zaretsky JM, et al. Mutations Associated with Acquired Resistance to PD-1 Blockade in Melanoma. *N Engl J Med*. 2016;375(9):819-829.
- Hänzelmann S, et al. GSVA: gene set variation analysis. *BMC Bioinformatics*. 2013;14:7.

## License

MIT
