---
title: "Introduction to irps"
author: "Maynard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to irps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `irps` package provides tools for evaluating PD-1/PD-L1 immunotherapy response potential from proteomics or transcriptomics expression data. The core metric, the **Immune Response Potential Score (IRPS)**, integrates information from 18 curated immune gene sets into a single interpretable scalar.

This vignette demonstrates the complete workflow:
1. Preparing input data
2. Computing IRPS
3. Decomposing the immune portrait
4. Classifying immune phenotypes
5. Validating against MSigDB Hallmarks
6. Generating publication-quality visualizations

## 1. Preparing Input Data

You need two inputs:
- **Expression matrix**: features (proteins or genes) × samples, normalized
- **Gene mapping table**: maps HGNC gene symbols to feature IDs in your expression matrix

```{r setup}
library(irps)

# Get the built-in immune gene sets
gs <- get_immune_gene_sets()
cat("Total gene sets:", length(gs), "\n")
cat("Gene set sizes:", lengths(gs), "\n")
```

### Simulated example data

```{r simdata}
set.seed(2024)

# Simulate 200 proteins × 50 samples
n_proteins <- 200
n_samples <- 50
protein_ids <- paste0("P", seq_len(n_proteins))
sample_ids <- paste0("S", seq_len(n_samples))

# Create expression matrix with some structure
expr_mat <- matrix(rnorm(n_proteins * n_samples, mean = 10, sd = 2),
                   nrow = n_proteins, ncol = n_samples)
rownames(expr_mat) <- protein_ids
colnames(expr_mat) <- sample_ids

# Create gene mapping: some of our 200 proteins map to immune genes
known_immune_genes <- unique(unlist(gs))
# Pick first 140 immune genes and assign them to our protein IDs
mapped_genes <- known_immune_genes[1:min(140, length(known_immune_genes))]
gene_map <- data.frame(
  Gene      = mapped_genes,
  FeatureID = protein_ids[seq_along(mapped_genes)],
  stringsAsFactors = FALSE
)

# Give some structure: inject B2M and HLA-A signal into specific samples
b2m_prot <- gene_map$FeatureID[gene_map$Gene == "B2M"]
hlaa_prot <- gene_map$FeatureID[gene_map$Gene == "HLA-A"]
if (length(b2m_prot) > 0) {
  expr_mat[b2m_prot, 1:15] <- expr_mat[b2m_prot, 1:15] + 3  # higher in first 15
}
if (length(hlaa_prot) > 0) {
  expr_mat[hlaa_prot, 1:15] <- expr_mat[hlaa_prot, 1:15] + 2.5
}

# Create synthetic group labels (e.g., subtypes)
groups <- setNames(
  c(rep("Subtype1", 15), rep("Subtype2", 18), rep("Subtype3", 17)),
  sample_ids
)

head(gene_map)
dim(expr_mat)
```

## 2. Computing IRPS

The `compute_irps()` function is the main entry point. It:
1. Maps gene symbols to feature IDs in your expression data
2. Runs ssGSEA for all available gene sets
3. Computes the IRPS composite score

```{r compute}
result <- compute_irps(
  expr_mat  = expr_mat,
  gene_map  = gene_map,
  platform  = "proteomics",
  groups    = groups,
  verbose   = FALSE
)

print(result)
```

## 3. Decomposing the Immune Portrait

The `decompose_portrait()` function breaks IRPS into 6 interpretable biological dimensions:

```{r portrait}
portrait <- decompose_portrait(result, expr_mat = expr_mat, gene_map = gene_map)
head(portrait[, 1:5])
```

## 4. Classifying Immune Phenotypes

```{r phenotype}
pheno <- classify_phenotype(result)
table(pheno$Immune_Phenotype, pheno$Subtype)
```

## 5. Visualization Gallery

### IRPS Boxplot

```{r boxplot, fig.width=5.5, fig.height=5}
plot(result, type = "boxplot")
```

### Immune Heatmap

```{r heatmap, fig.width=10, fig.height=6}
plot(result, type = "heatmap")
```

### Portrait Dimension Panel

```{r portrait_panel, fig.width=10, fig.height=7}
plot_portrait_panel(portrait)
```

### Immune Phenotype Distribution

```{r pheno_bar, fig.width=5.5, fig.height=4.5}
plot_phenotype_bar(pheno)
```

### IRPS vs Key Biomarker

```{r scatter, fig.width=6, fig.height=5}
if ("B2M" %in% gene_map$Gene) {
  plot_irps_scatter(result, expr_mat, gene_map, "B2M")
}
```

## 6. Validating Against MSigDB Hallmarks

*Note: requires the `msigdbr` package to be installed.*

```{r hallmark, eval=FALSE}
val <- validate_hallmark(result, expr_mat, gene_map)

# Inspect strongest correlations
cor_mat <- val$correlation_matrix
cat("Strongest correlation:", max(cor_mat, na.rm = TRUE), "\n")
```

## Session Information

```{r session}
sessionInfo()
```
